A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates

### 2024-02-13 paper/paper.qmd
update figure text positioning

---
title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
#subtitle: "My subtitle if needed"
author: 
  - Maroosh Gillani
  - Aaron Liu
  - Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
date: today
date-format: long
abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We explore how different social, cultural, and economic effects play a part in this worrying trend. This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
# workplace setup

@@

# replicate figure 2a from the original paper
fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
  geom_text(aes(x = 1985, y = 105), label = "Age 20-24", color = "black") +
  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
  theme_classic() +
  theme(legend.position = "none")

# replicate figure 2d from the original paper
fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Level)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother's level of education (ages 20-44)") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1995, y = 105), label = "No high school degree", color = "black") +
  geom_text(aes(x = 1995, y = 75), label = "College graduate
", color = "#d44215") +
  theme_classic() +
  geom_text(aes(x = 1995, y = 65), label = "High school graduate", color = "#0cc4be") +
  geom_text(aes(x = 1995, y = 45), label = "Some college", color = "black") +
  theme_classic() +
  theme(legend.position = "none")

# replicate figure 2f from the original paper
merged_data_long %>% ggplot(aes(x = year, y = round(Birth_Rate, digits = 1), color = Parity)) +
  geom_line(size = 1, stat = "summary", fun = mean) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Parity", caption = "Parity (ages 15-44)") +
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1995, y = 28), label = "First birth", color = "black") +
  geom_text(aes(x = 1995, y = 23), label = "Second birth", color = "black") +
  geom_text(aes(x = 1995, y = 12.5), label = "Third Birth", color = "black") +
  geom_text(aes(x = 1995, y = 8.5), label = "Fourth+ birth", color = "black") +
  theme_classic() +
  theme(legend.position = "none")
```

>>>>

---
title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
#subtitle: "My subtitle if needed"
author: 
  - Maroosh Gillani
  - Aaron Liu
  - Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
date: today
date-format: long
abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We explore how different social, cultural, and economic effects play a part in this worrying trend. This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
# workplace setup

@@

# replicate figure 2a from the original paper
fig_2a_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Age_Group)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Age Group", caption = "Five-year age group") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 5)) +
  geom_text(aes(x = 1985, y = 120), label = "Age 25-29", color = "black") +
  geom_text(aes(x = 1985, y = 100), label = "Age 20-24", color = "black") +
  geom_text(aes(x = 1985, y = 80), label = "Age 30-34", color = "black") +
  geom_text(aes(x = 1985, y = 58), label = "Age 15-19", color = "black") +
  geom_text(aes(x = 1985, y = 33), label = "Age 35-39", color = "black") +
  geom_text(aes(x = 1985, y = 10), label = "Age 40-44", color = "black") +
  theme_classic() +
  theme(legend.position = "none")

# replicate figure 2d from the original paper
fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Level)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Education Level", caption = "Mother's level of education (ages 20-44)") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1993, y = 110), label = "No high school degree", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
  theme_classic() +
  geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
  theme_classic() +
  theme(legend.position = "none")

# replicate figure 2f from the original paper
merged_data_long %>% ggplot(aes(x = year, y = round(Birth_Rate, digits = 1), color = Parity)) +
  geom_line(size = 1, stat = "summary", fun = mean) +
  labs(x = "Year", y = "Births per 1,000 women in relevant population subgroup", color = "Parity", caption = "Parity (ages 15-44)") +
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1993, y = 28), label = "First birth", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 23), label = "Second birth", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 12.5), label = "Third Birth", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 8.5), label = "Fourth+ birth", color = "black", hjust = 0) +
  theme_classic() +
  theme(legend.position = "none")
```
